|

### 2020-05-19 paper/Manuscript.Rmd
Update writing

Although the principles underlying this workflow are universal, WORCS has been optimized for R [@rcoreteamLanguageEnvironmentStatistical2020].
Several arguments support the choice to work in R.
First, R and all of it extensions are free and Open Source, and Open Source software should be the tool of choice for open science.
Second, as of this writing, R is the most advanced programming language in terms of the implementation of tools required for an open science workflow.
Third, R is the second-most cited statistical software package [@muenchenPopularityDataScience2012], following SPSS, which has no support for any of the open science tools discussed in this paper.
Fourth, R is well-supported; a vibrant online community exists for R, developing new R methods and packages, and providing support and tutorials for existing ones.
Finally, R is highly interoperable:
Packages are available to load nearly every imaginable filetype, and output can be written to most file types, including DOCX and PDF (in APA style), and HTML format. 
Moreover, wrappers are available for many tools developed in other programming languages, and code written in other programming languages can be evaluated from R [including C++, Fortran, and Python, @allaireMarkdownPythonEngine]. There are excellent free resources for learning R, such as the book R for Data Science [@grolemundDataScience2017].

Working with R is simplified immensely by using the RStudio integrated development engine (IDE) for R [@rstudioteamRStudioIntegratedDevelopment2015].
Most of the tools used in WORCS are embedded directly into the user interface of RStudio.
RStudio automates and streamlines tedious or complicated aspects of working with R, Rmarkdown, and Git/GitHub.
This makes RStudio a comprehensive solution for open science research projects.
Another important feature of RStudio is project management.
A project bundles writing, analyses, data, references, etcetera, into a self-contained folder, that can be uploaded entirely to GitHub and downloaded by future users.
The `worcs` R-package installs a new RStudio project template.
When a new project is initialized from this template, the bookkeeping required to set up an open science project is performed automatically.

@@

WORCS relies on several free, open source software solutions.
The first of these is *dynamic document generation*.
Dynamic document generation (DDG) means writing scientific reports in a format
that interleaves written reports with blocks of computer code used to conduct
the analyses. The text is automatically formatted as an APA-style manuscript
[thanks to the R-package `papaja`, @austPapajaPrepareReproducible2020], and
results of the code blocks are automatically generated and insterted in the
text, or rendered as Figures and Tables. Dynamic document generation supercedes
the classical approach of using separate programs to write prose and conduct
analyses, and then manually copy-pasting analysis results into the text.
Although there is a slight learning curve to transitioning to DDG, the
investment really pays off:

@@

The second solution is *version control*:
Maintaining an indelible log of every change to all project files. Version 
control is a near-essential tool for scientific reproducibility, as anyone
learns who has had the experience of accidentally deleting a crucial file, or of
being unable to reproduce analyses because they ran some analyses interactively
instead of using syntax [see also @blischakQuickIntroductionVersion2016].
Many scientists use some form of *implicit* version control; for example, by
renaming files after major changes (e.g., "manuscript_final_2.2-2019-10-12.doc"),
tracking changes in word processing software, or using cloud hosting services
that retain backups of previous versions.
WORCS instead uses the *explicit* version control software 'Git'
(www.git-scm.com). Git tracks
changes to files, and stores these changes when the user makes a "commit" (a 
snapshot of the version controlled files).
<!--Barbara: I have looked for a way to cite Git but did not find one... 
Seems good practice especially in this paper as an example to cite all software,
but for Git I just cannot find the way. Maybe cite the website: www.git-scm.com
CJ: Other papers discussing scientific merits of Git also just give a URL -->
Git retains a complete log of all commits, and users can compare changes between
different commits, or go back to a previous version of the code (for example,
after making a mistake, or to replicate a previous version of the results).
A project version controlled with Git is called a "Repository", or Repo.
Git only version controls files explicitly committed by the user. Moreover, it
is possible to prevent files from being version controlled - which is useful for
privacy sensitive data.
The `.gitignore` file lists files that should not be version-controlled.
These files thus exist only on the user's private computer.

@@

The third solution is *dependency management*: Keeping track of exactly what
software was used to conduct the analyses. At first glance, it might seem
sufficient to state that analyses were conducted in *Program X*. However, every
program is susceptible to changes, updates, and bugfixes. R-packages in
particular are updated frequently, because they are Open Source, and there is an
active community of developers contributing functionality and bugfixes. 
Potentially, any update could change the results of the code. For example, 
R-version 3.6.0 fixed a bug in the random number generator. This means that any
analyses that use random numbers, such as bootstrapping, cannot be strictly
replicated from previous versions. Or, for example, R-version 4.0.0 changed a
default parameter that affects how data are loaded from files.

<!-- LB: and onwards: this section requires some background knowledge on R/packages/etc. This could be a place where readers stumble and stop reading. Perhaps introduce the concept of the R project template earlier, to see section on dependencies more into perspective. 
[main issue] it should become clear earlier what the WORCS workflow entails: e.g. switching to working with a particular R Project Template. Discuss the ‘workflow' section earlier. Related: Can you make a schematic overview of the WORCS?overview--> 
Many solutions exist to ensure computational reproducibility. These solutions typically work by
enveloping your research project in a distinct "environment" that only has access to programs that
are explicitly installed, and maintaining a record of these programs. These solutions differ in
user-friendliness and effectiveness. If strict reproducibility is required, one might use the
aforementioned Docker-based workflow by @peikertReproducibleDataAnalysis2019, which runs all
analyses inside a "container": An environment that behaves like a virtual computer that can be
stored like a sort of time capsule, and identically reinstated on a user computer, or in the cloud.
This is effective, and also preserves software outside the R environment (e.g., preprocessing tools
for neuroimaging data). However, it can be difficult to set up for novice users - although a package
is in development to facilitate this transition [@repo]. The cloud-based research collaboration platform
'Code Ocean' is also based on Docker, for example. It is user-friendly, but it is not clear whether the
["Data Processing Agreement"](https://www.digitalocean.com/legal/data-processing-agreement/) is fully
GDPR compliant - it seems to focus more on limiting liability than on compliance. This is a risk with
any cloud-based solution. The advantage of using a local system is that data sharing is optional, and
WORCS capitalizes on this flexibility.

A slightly less sophisticated solution is offered by the R-package `packrat`, which installs all software
used in a project to a dedicated library that can be stored and copied with the project. This solution
is reasonably user-friendly and effective, but has the disadvantage that for each project, all packages
must be installed anew, which can take a long time (sometimes hours) and a lot of hard drive space.
Moreover, Git/GitHub is not designed to store your entire project library, so `packrat` does not address
the issue of how other users are supposed to get access to your library. In conclusion, `packrat` is
somewhat unwieldy. The package `checkpoint` offers an extremely lightweight solution: Instead of storing
all software used, it simply records the calendar date, and then installs the version of all required
packages that was available on the central R repository `CRAN` on that date from an archival website.
Thus, all that is needed to restore a project is the date on which it was conducted. This is very
user-friendly, and reasonably effective - the only problem is that this solution is limited to `CRAN`
packages, and cannot mix package versions from different dates.

A recently developed package strikes a great balance between user-friendliness and effectiveness: `renv`, developed by the team behind RStudio. This package maintains a text-based log file of which packages are used, in what versions, and where they were installed from (e.g., CRAN, Bioconductor, GitHub). Being text-based, this log can be version controlled with Git. On a local computer, `renv` installs all packages used in a cache that is shared between all projects that use `renv`. Thus, if one of your projects uses a version of a package that is already in the cache, it does not need to be installed again. This overcomes the limitations of long installation times and large space requirements of `renv`. When someone else loads your project, `renv` will install all of the required packages onto a cache on their computer. Whereas `packrat` copies the source code of all packages, `renv` obtains the packages from their original repositories. Thus, as long as these repositories used maintain their archives, this solution is very lightweight and effective. It is not foolproof, but some have argued that all solutions for computational reproducibility have a limited shelf life [@brownHowLearnedStop2017]. Ultimately, the best solution might be to use a good-enough solution, like `renv`, and acknowledge that all code requires some maintenance if you want to reproduce it in the future.

## Considering filetypes
A key consideration when developing a research project is what filetypes to use.
A case can be made to use text-based files only (or as much as possible), instead of binary files.
Text-based files can be read by machines and humans alike, without requiring special (licensed) software.
Binary files, such as Word (`.docx`) or SPSS (`.sav`, `.spo`),  must be decoded by software first.
Git is designed to keep track of changes to the content of text-based files.
Each change reflects a word or line of code added or removed.
Binary files, on the other hand, should not be version controlled with Git, as individual changes in the encrypted file are uninterpretable.
Moreover, text-based files are often smaller than binary files, which means they take up less space on cloud hosting services, such as GitHub. Uploading large binary files is frowned upon.
GitHub also allows visitors to peruse text-based files online, and conveniently renders certain filetypes.
For example, GitHub highlights syntax in `.R` and `.Rmd` files, and displays `.csv` files as spreadsheets.

>>>>

Although the principles underlying this workflow are universal, WORCS has been
designed for R [@rcoreteamLanguageEnvironmentStatistical2020].
Several arguments support the choice to work in R.
First, R and all of it extensions are free and open source software, which
should be the tool of choice for open science.
Second, as of this writing, R is the most advanced programming language in terms
of the implementation of tools required for an open science workflow.
Third, R is the second-most cited statistical software package [@muenchenPopularityDataScience2012], following SPSS, which has no support for any of the open science tools discussed in this paper.
Fourth, R is well-supported; a vibrant online community exists for R, developing new R methods and packages, and providing support and tutorials for existing ones.
Finally, R is highly interoperable:
Packages are available to load nearly every imaginable filetype, and output can be written to most file types, including DOCX and PDF (in APA style), and HTML format. 
Moreover, wrappers are available for many tools developed in other programming languages, and code written in other programming languages can be evaluated from R [including C++, Fortran, and Python, @allaireMarkdownPythonEngine]. There are excellent free resources for learning R, such as the book R for Data Science [@grolemundDataScience2017].

Working with R is simplified immensely by using the RStudio integrated development engine (IDE) for R [@rstudioteamRStudioIntegratedDevelopment2015].
Most of the tools used in WORCS are embedded directly into the user interface of RStudio.
RStudio automates and streamlines tedious or complicated aspects of working with R, Rmarkdown, and Git/GitHub.
This makes RStudio a comprehensive solution for open science research projects.
Another important feature of RStudio is project management.
A project bundles writing, analyses, data, references, etcetera, into a
self-contained folder, that can be uploaded entirely to 'GitHub' and downloaded
by future users. The `worcs` R-package installs a new RStudio project template.
When a new project is initialized from this template, the bookkeeping required
to set up an open science project is performed automatically. The `worcs` 
package has a [vignette](https://cjvanlissa.github.io/worcs/articles/setup.html)
on how to set up your system with all of the required tools.

@@

WORCS relies on several free, open source software solutions.
The first is *dynamic document generation* (DDG): Writing scientific reports in
a format that interleaves written reports with blocks of code used to conduct
the analyses. The text is automatically formatted; the `worcs` package can 
output manuscripts as an APA-style paper [thanks to the R-package `papaja`, @austPapajaPrepareReproducible2020], or as a host of other scientific formats 
[see @allaireRticlesArticleFormats2020], or as a plain 'Markdown' format for
'GitHub'. When the text is formatted, the results of the code blocks are
automatically generated and insterted in the text, or rendered as Figures and
Tables. Dynamic document generation supercedes
the classical approach of using separate programs to write prose and conduct
analyses, and then manually copy-pasting analysis results into the text.
Although there is a slight learning curve to transitioning to DDG, the
investment really pays off:

@@

The second solution is *version control*:
Maintaining an indelible log of every change to all project files. Version 
control is a near-essential tool for scientific reproducibility, as anyone
learns who has had the experience of accidentally deleting a crucial file, or of
being unable to reproduce analyses because they ran some analyses interactively
instead of using syntax [see also @blischakQuickIntroductionVersion2016].
Many scientists use some form of *implicit* version control; for example, by
renaming files after major changes (e.g., "manuscript_final_2.2-2019-10-12.doc"),
tracking changes in word processing software, or using cloud hosting services
that retain backups of previous versions.
WORCS instead uses the *explicit* version control software 'Git'
(www.git-scm.com). Git tracks
changes to files, and stores these changes when the user makes a "commit" (a 
snapshot of the version controlled files).
Git retains a complete log of all commits, and users can compare changes between
different commits, or go back to a previous version of the code (for example,
after making a mistake, or to replicate a previous version of the results).
A project version controlled with Git is called a "Repository", or Repo.
Git only version controls files explicitly committed by the user. Moreover, it
is possible to prevent files from being version controlled - which is useful for
privacy sensitive data.
The `.gitignore` file lists files that should not be version-controlled.
These files thus exist only on the user's private computer.

@@

The third solution is *dependency management*: Keeping track of exactly what
software was used to conduct the analyses. At first glance, it might seem
sufficient to state that analyses were conducted in *Program X*. However, every
program is susceptible to changes, updates, and bugfixes. R-packages in
particular are updated frequently, because they are Open Source, and there is an
active community of developers contributing functionality and bugfixes. 
Potentially, any update could change the results of the code. <!--For example, 
R-version 3.6.0 fixed a bug in the random number generator. This means that any
analyses that use random numbers, such as bootstrapping, cannot be strictly
replicated from previous versions. Or, for example, R-version 4.0.0 changed a
default parameter that affects how data are loaded from files.-->

<!-- LB: and onwards: this section requires some background knowledge on R/packages/etc. This could be a place where readers stumble and stop reading. Perhaps introduce the concept of the R project template earlier, to see section on dependencies more into perspective. 
[main issue] it should become clear earlier what the WORCS workflow entails: e.g. switching to working with a particular R Project Template. Discuss the ‘workflow' section earlier. Related: Can you make a schematic overview of the WORCS?overview--> 
Many solutions exist to ensure computational reproducibility. These solutions typically work by
enveloping your research project in a distinct "environment" that only has access to programs that
are explicitly installed, and maintaining a record of these programs. These solutions differ in
user-friendliness and effectiveness. If strict reproducibility is required, one might use the
aforementioned Docker-based workflow by @peikertReproducibleDataAnalysis2019, which runs all
analyses inside a "container": An environment that behaves like a virtual computer that can be
stored like a sort of time capsule, and identically reinstated on a user computer, or in the cloud.
This is effective, and also preserves software outside the R environment (e.g., preprocessing tools
for neuroimaging data). However, it can be difficult to set up for novice users - although a package
is in development to facilitate this transition [@repo]. The cloud-based research collaboration platform
'Code Ocean' is also based on Docker. It is very user-friendly, but it is not clear whether the
["Data Processing Agreement"](https://www.digitalocean.com/legal/data-processing-agreement/) is fully
GDPR compliant - it seems to focus more on limiting liability than on compliance. This is a risk with
any cloud-computing solution.

For analyses conducted entirely in R, a slightly less sophisticated solution is offered by the R-package `packrat`, which installs all packages
used in a project to a dedicated library that can be stored and copied with the project. This has the disadvantage that for each project, all packages
must be installed anew, which can take a long time (sometimes hours) and a lot of hard drive space.
Moreover, 'GitHub' is not designed to store your entire project library, so `packrat` does not address
the issue of how other users are supposed to get access to your library. In conclusion, `packrat` is
somewhat unwieldy. The package `checkpoint` offers an extremely lightweight solution: Instead of storing
all R-packages used, it simply records the calendar date, and when another user loads your project, it installs the version of all required
packages that was available on the central R repository `CRAN` on that date from an archival website.
Thus, all that is needed to restore a project is the date on which it was conducted. The main limitation is that this solution is limited to `CRAN`
packages, and cannot mix package versions from different dates.

A recently developed package strikes a great balance between user-friendliness 
and effectiveness: `renv`, developed by the team behind 'RStudio'. This package
maintains a text-based, human-readable log of the packages used, their version
numbers, and where they were installed from (e.g., 'CRAN', 'Bioconductor', 'GitHub').
This text-based log can be version controlled with 'Git'. `renv` installs all
packages in a cache that is shared between all projects that use `renv`. Thus,
if one of your projects uses a package that is already in the cache, it does not
need to be installed again. This overcomes the limitations of long installation
times and large space requirements of `packrat`. When someone else loads your 
project, `renv` will install all of the required packages onto a cache on their
computer. Whereas `packrat` copies the source code of all packages, `renv`
obtains the packages from their original repositories. Thus, as long as these
repositories are maintained, this solution is very lightweight and effective.
It is not foolproof, but all solutions for computational
reproducibility have a limited shelf life [@brownHowLearnedStop2017].
Ultimately, the best solution might be to use a good-enough solution, like
`renv`, and acknowledge that all code requires some maintenance if you want to
reproduce it in the future.

## Text-based files are better

A key consideration when developing a research project is what filetypes to use.
A case can be made to use text-based files only (or as much as possible),
instead of binary files.
Text-based files can be read by machines and humans alike.
Binary files, such as 'Word' (`.docx`) or 'SPSS' (`.sav`, `.spo`) files,
must be decoded by (commercial) software first.
'Git' is designed to track any changes to the content of text-based files.
Each change reflects a line of text added or removed. This means that, when
writing a version-controlled paper in plain-text, it is recommended to insert a
line break after every sentence (instead of after every paragraph). That way,
the change log will indicate which specific sentence was edited.
Binary files should not be version controlled with 'Git',
as changes in the encrypted file are uninterpretable. Binary files are often 
also larger than text-based files, which means they take up more space on cloud
hosting services, such as 'GitHub'. For those reasons, uploading large binary
files to these platforms is frowned upon.
'GitHub' does allow visitors to peruse text-based files online, and
conveniently renders certain filetypes: For example, 'GitHub' displays `.md`
(Markdown) files as web pages, and `.csv` files as spreadsheets. This is why
`worcs` uses Markdown and `.csv` extensively.
